A framework for pipeline benchmarking and its application to single-cell RNAseq clustering

### 2020-01-30 bmc_article.tex
Updates from Overleaf

\author[
   addressref={aff1,aff2},          % id's of addresses, e.g. {aff1,aff2}
   email={pierre-luc.germain@hest.ethz.ch}   % email address
]{\inits{PLG}\fnm{Pierre-Luc} \snm{Germain}}
\author[
   addressref={aff1},
    email={anthony.sonrel@uzh.ch}
]{\fnm{Anthony} \snm{Sonrel}}
\author[
   addressref={aff1},
   corref={aff1},                   % id of corresponding address, if any
   email={mark.robinson@imls.uzh.ch}
]{\inits{MR}\fnm{Mark} \snm{Robinson}}

@@

\address[id=aff1]{
  \orgname{Statistical Bioinformatics, IMLS, University of Z\"{u}rich}, % university, etc
  \street{Winterthurerstrasse 190},
  \postcode{8057}
  \city{Z\"{u}erich},
  \cny{Switzerland}
}
\address[id=aff2]{
  \orgname{D-HEST Institute for Neurosciences, ETH Z\"{u}rich},
  \street{Winterthurerstrasse 190},
  \postcode{8057}
  \city{Z\"{u}erich},
  \cny{Switzerland}
}

@@

\begin{abstract} % abstract
%\parttitle{Background} %if any
The massive growth of single-cell RNA sequencing (scRNAseq) and methods for its analysis is currently only insufficiently accompanied by proper and up-to-date benchmarks that would guide analytical choices. Moreover, current studies are often focused on isolated steps of the process. Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics, and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities. We evaluate common steps of such analyses, including filtering, doublet detection (suggesting a new R package, \textit{scDblFinder}), normalization, feature selection, denoising, dimensionality reduction and clustering. On the basis of these analyses, we make a number of key concrete recommendations about analysis choices. The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool and permit further extending benchmarks (\url{https://github.com/plger/pipeComp}).

@@

Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, with more than 560 software tools available to the community \cite{ZappiaDB2018}, roughly half of which are dedicated to tasks related to data processing such as clustering, ordering, dimension reduction or normalization. This increase in the number of available tools follows the development of new sequencing technologies and the growing number of reported cells, genes and cell populations \cite{SvenssonDB2019}. Data processing is a critical step in any scRNAseq analysis, affecting downstream analysis and interpretation. It is therefore critical to evaluate the available tools to leverage future studies as it can ultimately increase the power of further differential expression analysis \cite{viethSystematic2019}.

@@

Here, we develop a flexible R framework for pipeline comparison and evaluate in details the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We harnessed real datasets of known cell composition (Table \ref{tab:table1}) and develop new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}). 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly. The \textit{CellBench} package was recently proposed to address a similar need\citep{su_cellbench}, offering an elegant piping syntax to combine alternative methods at successive steps. In addition to offering the same capabilities, a key additional feature of \textit{pipeComp} is that, since the evaluation metrics are not treated as a step but parallel to it, they can be performed at various points in the pipeline without needing to store the potentially large intermediate data, and thereby enabling combinatorial benchmarks. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI, \citep{HubertARI1985}) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and might not always capture improvements in earlier steps of the pipeline.

Another motivation for the present framework is that clustering is relatively fragile to variations in the pipeline. We therefore also wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters. We therefore proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (\ref{fig:figure2}B).

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \citep{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \citep{tianMixology2018}, and for which therefore SNP information can be used as ground truth. We included a further such dataset\citep{kangMultiplexedDemuxlet2018} not included for the rest of our benchmark due to the absence of true cell labels beside SNPs. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals), whose identification is generally not the primary aim of doublet detection methods, since these can be considered homotypic (as opposed to neotypic or heterotypic doublets, i.e. doublets from different cell types) and can be identified through other means.

We tested \textit{DoubletFinder} \citep{mcginnisDoubletfinder2019} and \textit{scran}'s \texttt{doubletCells} \citep{LunScran2016}, both of which use similarity to artificial doublets, and \textit{scds} \citep{BaisScds2019}, which relies on a combination of co-expression and binary classification. \textit{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \textit{scran} and \textit{scds} return scores that must be manually thresholded. In these cases, we ensured that the right number of cells would be called doublets. In addition to these methods, we reasoned that an approach such as \textit{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets more efficiently. We therefore developed a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \citep{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function of \textit{scater} that measures, for a given control property, the number of median absolute deviations of each cell from the median of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.

@@

We next investigated the impact of filtering according to various criteria (see methods). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g. library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of median absolute deviations (MADs) from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 8). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here, and we therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \citep{PapadimitriouHu1998}, and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious, and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias. Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out. This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy. We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see methods).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see methods for details).}
    \label{fig:figure4}
\end{figure}

@@

An important motivation for the development \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}. We therefore first assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A). The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \textit{Seurat} normalization resulted in fairly high correlation with technical covariates. \textit{sctransform} led to the lowest correlation but most methods, including normalization based on stable genes, were able to remove most of the correlation when combined with scaling. The only exception is \textit{Seurat} normalization with scaling regressing out covariates, which surprisingly increased the correlation in 8 of the 9 datasets.

We next investigated the impact of normalization on the separability of the subpopulations (Figure \ref{fig:figure5}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width\citep{RousseeuwSil1987} and mutual information. We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones, and was generally but not always beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave poorer performance on all metrics. \textit{sctransform} systematically outperformed other methods, and even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with Smart-seq protocol (Koh and Kumar datasets).

@@

A standard clustering pipeline typically involves a step of highly-variable genes selection, which is complicated by the digital nature and the mean-variance relationship of (sc)RNAseq. \textit{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions (>3) rely on a different measure of variance, again standardized. Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this relationship may in fact sometimes reflects biological relevance and would be helpful in classifying cell types. Another common practice in features selection is to use those with the highest mean expression, and recently, \citep{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three such alternatives. Analysis of variance performed on a standard \textit{Seurat} normalization or on \textit{sctransform} data were highly correlated (Supplementary Figure 11A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 12-14). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding to the percentage of deviance explained however, the standardized measures were often superior(Figure \ref{fig:figure6}A and Supplementary Figures 12-13). Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.

@@

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B). To validate the previous assay on the proportion of variance/ deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The previous observations were reflected in the ARI of the resulting clustering (Figure \ref{fig:figure6}B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 14).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (e.g. Supplementary Figure 15). The first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations. Seurat's PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7}). 
Overall, weighting the principal components by their variance in the same fashion as Seurat's method does had a strong positive impact on silhouette widths and ARI scores.

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 15). This tended to be translated to the accuracy of the clustering (Supplementary Figure 16), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.

@@

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the walktrap method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textitt{scran} methods (walktrap and fast greedy) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, while \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified, and therefore could be used to find a parameter combination producing the right number of clusters, the number of clusters found by \textit{scran}-based clustering was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 17). The occasional lack of scran-based clustering with the right number of clusters, along with a higher mutual information, suggest that \textit{scran} sometimes simply divides a real subpopulation into two (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the walktrap method appeared superior to the fast greedy algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. Of note, very difficult subpopulations, both from the Zhengmix8eq and simMix1 datasets, remained very inaccurately classified by all methods and in regards to all metrics.

>>>>

\author[
   addressref={aff1,aff2},          % id's of addresses, e.g. {aff1,aff2}
   email={pierre-luc.germain@hest.ethz.ch}   % email address
]{\inits{PLG}\fnm{Pierre-Luc} \snm{Germain}}
\author[
   addressref={aff1},
    email={anthony.sonrel@uzh.ch}
]{\fnm{Anthony} \snm{Sonrel}}
\author[
   addressref={aff1},
   corref={aff1},  % id of corresponding address, if any
   email={mark.robinson@imls.uzh.ch}
]{\inits{MR}\fnm{Mark D.} \snm{Robinson}}

@@

\address[id=aff1]{
  \orgname{Department of Molecular Life Sciences and SIB Swiss Institute of Bioinformatics, University of Zürich}, % university, etc
  \street{Winterthurerstrasse 190},
  \postcode{8057}
  \city{Zürich},
  \cny{Switzerland}
}
\address[id=aff2]{
  \orgname{D-HEST Institute for Neurosciences, ETH Zürich},
  \street{Winterthurerstrasse 190},
  \postcode{8057}
  \city{Zürich},
  \cny{Switzerland}
}

@@

\begin{abstract} % abstract
%\parttitle{Background} %if any
The massive growth of single-cell RNA sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices. Moreover, current studies are often focused on isolated steps of the process. Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics, and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities. We evaluate common steps of such analyses, including filtering, doublet detection (suggesting a new R package, \textit{scDblFinder}), normalization, feature selection, denoising, dimensionality reduction and clustering. On the basis of these analyses, we make a number of concrete recommendations about analysis choices. The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool and permit extensible benchmarks (\url{https://github.com/plger/pipeComp}).

@@

Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, with more than 560 software tools available to the community \cite{ZappiaDB2018}, roughly half of which are dedicated to tasks related to data processing such as clustering, ordering, dimension reduction or normalization. This increase in the number of available tools follows the development of new sequencing technologies and the growing number of reported cells, genes and cell populations \cite{SvenssonDB2019}. As data processing is a critical step in any scRNAseq analysis, affecting downstream analysis and interpretation, it is critical to evaluate the available tools.

@@

Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We harnessed real datasets of known cell composition (Table \ref{tab:table1}) and develop new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for a description of the datasets. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}). 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly. The \textit{CellBench} package was recently proposed to address a similar need\citep{su_cellbench}, offering an elegant piping syntax to combine alternative methods at successive steps. In addition to offering the same capabilities, a key additional feature of \textit{pipeComp} is that, since the evaluation metrics are not treated as a step but parallel to it, they can be performed at various points in the pipeline without needing to store the potentially large intermediate data, and thereby enabling combinatorial benchmarks. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI, \citep{HubertARI1985}) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and might not always capture improvements in earlier steps of the pipeline.

Another motivation for the present framework is that clustering is relatively fragile to variations in the pipeline. We therefore also wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters. We therefore proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \citep{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \citep{tianMixology2018}, using SNP genotypes as the ground truth. For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \citep{kangMultiplexedDemuxlet2018}. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals), whose identification is generally not the primary aim of doublet detection methods, since these can be considered homotypic (as opposed to neotypic or heterotypic doublets, i.e. doublets from different cell types) and can be identified through other means.

We tested \textit{DoubletFinder} \citep{mcginnisDoubletfinder2019} and \textit{scran}'s \texttt{doubletCells} \citep{LunScran2016}, both of which use similarity to artificial doublets, and \textit{scds} \citep{BaisScds2019}, which relies on a combination of co-expression and binary classification. \textit{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \textit{scran} and \textit{scds} return scores that must be manually thresholded. In these cases, we ensured that the right number of cells would be called doublets. In addition to these methods, we reasoned that an approach such as \textit{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets more efficiently. We therefore developed a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see Methods).

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \citep{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.

@@

We next investigated the impact of filtering according to various criteria (see Methods). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g. library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 8). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here, and we therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \citep{PapadimitriouHu1998}, and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious, and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out. This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy. We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
    \label{fig:figure4}
\end{figure}

@@

An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}. We therefore first assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A). The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \textit{Seurat} normalization resulted in fairly high correlation with technical covariates. \textit{sctransform} led to the lowest correlation but most methods, including normalization based on stable genes, were able to remove most of the correlation when combined with scaling. The only exception is \textit{Seurat} normalization with scaling regressing out covariates, which surprisingly increased the correlation in 8 of the 9 datasets.

We next investigated the impact of normalization on the separability of the subpopulations (Figure \ref{fig:figure5}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \citep{RousseeuwSil1987} and mutual information. We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones, and was generally but not always beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave poorer performance on all metrics. \textit{sctransform} systematically outperformed other methods, and even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

@@

A standard clustering pipeline typically involves a step of highly-variable genes selection, which is complicated by the digital nature and the mean-variance relationship of (sc)RNAseq. \textit{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions ($\geq$3.0) rely on a different measure of variance, again standardized. Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this relationship may in fact sometimes reflects biological relevance and would be helpful in classifying cell types. Another common practice in feature selection is to use those with the highest mean expression, and recently, \citep{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three such alternatives. Analysis of variance performed on a standard \textit{Seurat} normalization or on \textit{sctransform} data were highly correlated (Supplementary Figure 11A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 12-14). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior(Figure \ref{fig:figure6}A and Supplementary Figures 12-13). Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.

@@

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The previous observations were reflected in the ARI of the resulting clustering (Figure \ref{fig:figure6}B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 14).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (e.g. Supplementary Figure 15). The first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations. Seurat's PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7}). 
Overall, weighting the principal components by their variance, in the same fashion as Seurat's method, had a strong positive impact on silhouette widths and ARI scores.

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 15). This tended to be translated to the accuracy of the clustering (Supplementary Figure 16), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.

@@

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textitt{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, while \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified, and therefore could be used to find a parameter combination producing the right number of clusters, the number of clusters found by \textit{scran}-based clustering was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 17). The occasional lack of scran-based clustering with the right number of clusters, along with a higher mutual information, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. Of note, very difficult subpopulations, both from the Zhengmix8eq and simMix1 datasets, remained very inaccurately classified by all methods and in regards to all metrics.
